Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.

### 2023-04-12 scripts/manuscript/manuscript.tex
return from australia

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\usepackage{float}
\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother
\journal{Remote Sensing of Environment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.},
  pdfauthor={Evan R. Muise; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Margaret E. Andrew; Stephen S. Ban},
  pdfkeywords={structural equation modelling, remote
sensing, landsat, forest structure, forest productivity, dynamic habitat
indices},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Disentangling linkages between satellite derived forest structure
and productivity across a range of forest types and ecosystems.}
\author[1]{Evan R. Muise%
\corref{cor1}%
}
 \ead{evan.muise@student.ubc.ca} 
\author[1]{Nicholas C. Coops%
%
}
 \ead{nicholas.coops@ubc.ca} 
\author[2]{Txomin Hermosilla%
%
}
 \ead{txomin.hermosillagomez@NRCan-RNCan.gc.ca} 
\author[1]{A. Cole Burton%
%
}
 \ead{cole.burton@ubc.ca} 
\author[3]{Margaret E. Andrew%
%
}
 \ead{M.Andrew@murdoch.edu.au} 
\author[4]{Stephen S. Ban%
%
}
 \ead{Stephen.Ban@gov.bc.ca}

\affiliation[1]{organization={University of British Columbia, Forest
Resources Management},addressline={2424 Main Mall},city={Vancouver, BC,
Canada},postcode={V6T 1Z4},postcodesep={}}
\affiliation[2]{organization={Natural Resources Canada, Canada Forest
Service (Pacific Forestry Centre)},addressline={506 Burnside Rd
W},city={Victoria, BC, Canada},postcode={V8Z 1M5},postcodesep={}}
\affiliation[3]{organization={Murdoch University, Environmental and
Conservation Sciences and Harry Butler Institute},city={Murdoch, WA,
Australia},postcode={6150},postcodesep={}}
\affiliation[4]{organization={Ministry of Environment and Climate Change
Strategy, BC Parks},addressline={525 Superior Street},city={Victoria,
BC, Canada},postcode={V8V 1T7},postcodesep={}}

@@

\begin{keyword}
    structural equation modelling \sep remote
sensing \sep landsat \sep forest structure \sep forest
productivity \sep 
    dynamic habitat indices
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

Biodiversity is the summation of variation in biological life, across
genes, species, communities, and ecosystems. Currently, biodiversity is
in decline, facing extinction rates above the background extinction rate
\citep{thomas2004, urban2015}, and homogenization of communities at
various scales \citep{mcgill2015}. In response, the global biodiversity
community is making efforts to assess and halt the degradation of
biodiversity on earth. The Group for Earth Observation Biodiversity
Observation Network has developed the Essential Biodiversity Variables
\citep[EBVs,][]{pereira2013}, designed as an analog to the Essential
Climate Variables framework \citep{bojinski2014}. EBVs are designed to
be global in scope, relevant to biodiversity information, feasible to
utilize, and complementary to one another \citep{skidmore2021}.

There are six EBV classes, each corresponding to a different facet of
biodiversity, including, genetic composition, species populations,
species traits, community composition, ecosystem structure, and
ecosystem function \citep{pereira2013}. \citet{fernández2020} divides
the six classes into two approaches, with one focusing on species
biodiversity, and the other focusing on ecosystem diversity. Remote
sensing has proven to be capable of measure five of the six classes,
with the exception being genetic composition, which requires in-situ
observation and samples \citep{skidmore2021}. Notably, while remote
sensing can provide information on the remaining two species EBV
classes, this information is typically acquired with an \emph{ad-hoc}
approach requiring high spatial and spectral resolution data. This data
can often only be collected at local extents, rather than the global or
regional scales required for biodiversity trend assessment
\citep{valdez2023}. Community composition falls into a similar dilemma,
requiring species population information which necessitates high
resolution spatial data.

The remaining two classes, ecosystem structure and function, are
incredibly well suited to be examined at global or regional scales using
mid-resolution satellite imagery, such as that provided by the Landsat
series of satellites. Advances in satellite remote sensing processing
have allowed 3d forest structure data to be imputed across wide spatial
scales \citep{matasci2018, coops2021} using data fusion approaches
involving collected lidar data and optical/radar data. Other advances in
image compositing have allowed yearly summaries of vegetation
productivity to be calculated at regional to global scales, summarizing
the yearly energy totals, minimums, and variations \citep{radeloff2019}.
These datasets correspond quite well with the EBV classes ecosystem
structure (forest structural diversity metrics), and ecosystem function
(forest productivity metrics).

Forest structural diversity has been linked to biodiversity at various
scales \citep{guo2017, bergen2009, gao2014}. Increased structural
complexity is hypothesized to create additional niches, leading to
increased species diversity \citep{bergen2009}. The relationship between
forest structure and biodiversity is commonly assessed using avian
species diversity metrics \citep{macarthur1961, goetz2007}, however,
other clades (and habitats), have also been used
\citep{davies2014, nelson2005}. Many metrics derived from lidar remote
sensing have been used as local indicators of biodiversity, including
canopy cover, canopy height, vertical profiles, and aboveground biomass,
while other 2nd order derived metrics such as canopy texture, height
class distribution, edges, and patch metrics have been used to examine
habitat and biodiversity at landscape scales \citep{bergen2009}.

@@

Some vegetation structure metrics are simpler, and more accurate, to
calculate than others \citep{coops2021}. These basic metrics, such as
canopy height and canopy cover, can then be used to estimate additional
structure metrics, such as basal area or total biomass. Canopy height
and cover are commonly used as indicators of vertical and horizontal
variation, respectively. Recently, more attention has been paid to
internal structural complexity metrics, which can be more difficult and
time consuming to generate \citep{coops2021, ma2022}, but have been
shown to have stronger linkages with biodiversity \citep{guo2017}, and
productivity \citep{ali2019}.

Linkages between forest structure and productivity (namely, vegetation
indices) have been examined for nearly 20 years \citetext{\citealp[
\citet{knyazikhin1998}]{huete2002}; \citealp{myneni1994}}. While there
is significant theoretical and empirical evidence for their relationship
at single time points (within a single image) \citep{myneni1994},
various relationship directions and shapes have been found between
forest structure and productivity metrics \citep{ali2019}. These
relationships, their shapes, and their strengths have been attributed to
multiple possible hypotheses and have been shown to vary based on
environmental conditions \citep{ali2019}. The relationship between
forest structural diversity metrics and annual productivity summaries
has yet to be fully examined, including the DHIs.

Understanding the complementarity between potential EBVs is an integral
component of their creation \citep{skidmore2021}. Biodiversity has been
shown to be linked to both forest structure \citep{guo2017, gao2014},
and productivity \citep{radeloff2019} at a range of scales. Research has
highlighted that the relationship between productivity and biodiversity
is reciprocal \citep{worm2003}, making productivity a suitable
biodiversity indicator. In forested ecosystems, stand structural
attributes have also been shown to influence productivity, with a range
of responses dependent on environmental conditions \citep{ali2019}.

Structural equation modelling and path analyses have been commonly used
in ecology to assess the causal effects behind various hypotheses
\citep{fan2016, grace2010}, including the relationship between forest
structure, functioning, and biodiversity \citep{ali2019} . In this paper
we seek to untangle the relationship between two EBVs: forest structure
diversity metrics and yearly summaries of forest productivity. To
accomplish this, we assess this relationship using path analysis to
assess the direct and indirect (as mediated by more complex forest
structural diversity metrics) effects of commonly collected forest
structural metrics (canopy height and canopy cover) on yearly
productivity summaries. Further, we use exploratory structural equation
modelling (ESEM) to identify the latent variables driving the
productivity metrics. We run this analysis both globally, and stratified
by the forested ecosystems of British Columbia, Canada, and compare
models using AIC scores and global fit measures across the ecosystems.
The results from this study will assess the linkages and complementarity
between two biodiversity indicators.

@@

We conducted model based sampling across the fifteen forest dominated
ecosystems found within British Columbia {[}BC figure reference, I have
not made this figure yet{]}. Samples were randomly selected within BEC
zones of interest alongside multiple criteria. Each sampled pixel had to
have a forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class.
Additionally, each pixel had to have a coefficient of variation less
than 0.5 in the Lorey's height and canopy cover forest structural
metrics. These focal analyses were preprocessed in R and Python, and
randomly sampled across a masked suitability raster for each BEC zone
using the \emph{sgsR} R package version 1.3.4 \citep{goodbody2023}. A
maximum of 3000 samples were taken from each BEC zone.

{[}note to ncc; I've stopped doing this sampling step as Meg said she
didn't feel it was necessary{]}

The imputation method used to generate the forest structure layers leads
to many duplicated values as there are a limited number of samples to
predict from \citep{matasci2018}. A discussion of imputation approaches
to large area lidar attribute modelling can be found in
\citet{coops2021}. To account for duplicated input values in our
modelling we averaged the endogenous variables within each ecosystem for
each uniquely imputed set of forest attributes, forest type, and
ecosystem. Due to variations in ecosystem size, differing numbers of
samples were generated for each ecosystem. To account for this during
model comparison, each ecosystem and land cover sample set was
downsampled to the minimum number of samples for a single BEC zone (in
this case, 170). For the forest type models, we attempted to select
equal numbers of observations from each BEC zone. However, some forest
types are rarely found in certain zones. To account for this, if an
equal number of samples from each BEC zone were not available, the
entire sample from that zone was taken. The amount of samples taken from
other zones was then increased in order to collect a total number equal
to the minimum number of samples in a single BEC zones (170). To meet
the normality assumptions of path analysis, all variables were
natural-log transformed and standardized, as per \citet{grace2016}.
Variables containing zeros were natural-log plus one transformed.

@@

@@

{\centering \includegraphics[width=14.74in,height=\textheight]{../../outputs/latent_groups.png}

@@

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

>>>>

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\usepackage{float}
\makeatletter
\let\oldlt\longtable
\let\endoldlt\endlongtable
\def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
\def\longtable@i[#1]{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt[#1]
}
\def\longtable@ii{\begin{figure}[H]
\onecolumn
\begin{minipage}{0.5\textwidth}
\oldlt
}
\def\endlongtable{\endoldlt
\end{minipage}
\twocolumn
\end{figure}}
\makeatother
\journal{Remote Sensing of Environment}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Disentangling linkages between satellite derived forest structure and productivity across a range of forest types and ecosystems.},
  pdfauthor={Evan R. Muise; Margaret E. Andrew; Nicholas C. Coops; Txomin Hermosilla; A. Cole Burton; Stephen S. Ban},
  pdfkeywords={structural equation modelling, remote
sensing, landsat, forest structure, forest productivity, dynamic habitat
indices},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Disentangling linkages between satellite derived forest structure
and productivity across a range of forest types and ecosystems.}
\author[1]{Evan R. Muise%
\corref{cor1}%
}
 \ead{evan.muise@student.ubc.ca} 
\author[2]{Margaret E. Andrew%
%
}
 \ead{M.Andrew@murdoch.edu.au} 
\author[1]{Nicholas C. Coops%
%
}
 \ead{nicholas.coops@ubc.ca} 
\author[3]{Txomin Hermosilla%
%
}
 \ead{txomin.hermosillagomez@NRCan-RNCan.gc.ca} 
\author[1]{A. Cole Burton%
%
}
 \ead{cole.burton@ubc.ca} 
\author[4]{Stephen S. Ban%
%
}
 \ead{Stephen.Ban@gov.bc.ca}

\affiliation[1]{organization={University of British Columbia, Forest
Resources Management},addressline={2424 Main Mall},city={Vancouver, BC,
Canada},postcode={V6T 1Z4},postcodesep={}}
\affiliation[2]{organization={Murdoch University, Environmental and
Conservation Sciences and Harry Butler Institute},city={Murdoch, WA,
Australia},postcode={6150},postcodesep={}}
\affiliation[3]{organization={Natural Resources Canada, Canada Forest
Service (Pacific Forestry Centre)},addressline={506 Burnside Rd
W},city={Victoria, BC, Canada},postcode={V8Z 1M5},postcodesep={}}
\affiliation[4]{organization={Ministry of Environment and Climate Change
Strategy, BC Parks},addressline={525 Superior Street},city={Victoria,
BC, Canada},postcode={V8V 1T7},postcodesep={}}

@@

\begin{keyword}
    structural equation modelling \sep remote
sensing \sep landsat \sep forest structure \sep forest
productivity \sep 
    dynamic habitat indices
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi

\begin{verbatim}
This is lavaan 0.6-14
lavaan is FREE software! Please report any bugs.
\end{verbatim}

\begin{verbatim}
terra 1.7.3
\end{verbatim}

\begin{verbatim}
Linking to GEOS 3.9.3, GDAL 3.5.2, PROJ 8.2.1; sf_use_s2() is TRUE
\end{verbatim}

@@

Biodiversity is the summation of variation in biological life, across
genes, species, communities, and ecosystems. Currently, biodiversity is
in decline, facing extinction rates above the background extinction rate
\citep{thomas2004, urban2015}, and homogenization of communities at
various scales \citep{mcgill2015}. In response, the global biodiversity
community is making efforts to assess and halt the degradation of
biodiversity on earth. The Group for Earth Observation Biodiversity
Observation Network has developed the Essential Biodiversity Variables
\citep[EBVs,][]{pereira2013}, designed as an analog to the Essential
Climate Variables framework \citep{bojinski2014}. EBVs are designed to
be global in scope, relevant to biodiversity information, feasible to
utilize, and complementary to one another \citep{skidmore2021}. There
are six EBV classes, each of which correspond to a different facet of
biodiversity including species populations, species traits, community
composition, ecosystem structure, ecosystem function, and genetic
composition \citep{pereira2013}.

Satellite remote sensing has proven to be capable of measuring five of
the six classes, with the exception being genetic composition, which
requires in-situ observation and sampling \citep{skidmore2021}. Species
populations - and in turn community composition - can be assessed with
very-high-resolution imagery to identify tree species at the tree-crown
scale, however it is difficult and computationally expensive to extend
these analyses to broader extents \citep{fassnacht2016, graves2016}.
Species traits such as phenology have been on a single-tree scale using
PlanetScope imagery and drone-based measurements \citep{wu2021}. Coarser
measurements, such as those taken by the MODIS \citep{zhang2003},
Landsat \citep{fisher2006}, or Sentinel
\citep{helfenstein2022, darvishzadeh2019} series of satellites can
monitor these functional processes at larger extents, corresponding to
ecosystem function, but the coarse spatial resolution removes the
ability to relate these traits to individual trees. Remote sensing data
has been shown to be effective at monitoring ecosystem structure at
regional to global extents through the use of optical imagery
\citep{cohen2004} and lidar datasets
\citep{lefsky2002, lang2021, neuenschwander2019}.

Notably, while remote sensing can provide information on the
species-level EBVs, this information is typically acquired with an
\emph{ad-hoc} approach requiring high spatial and spectral resolution
data \citep{skidmore2021}. This data can often only be collected at
small extents, rather than the global or regional scales required for
biodiversity trend assessment \citep{valdez2023}. The two
landscape-level EBVs (ecosystem structure and function) are well suited
to be examined at global or regional scales using moderate-resolution
satellite imagery, such as those provided by the Landsat or Sentinel
series of satellites \citep{skidmore2021}.

Simple metrics to complex for structure.

Forest structural diversity has been linked to biodiversity at various
scales \citep{guo2017, bergen2009, gao2014}. Increased structural
complexity is hypothesized to create additional niches, leading to
increased species diversity \citep{bergen2009}, which has been
frequently demonstrated using avian species diversity metrics
\citep{macarthur1961}. For example: \citet{herniman2020} used spectral
and lidar derived forest structure data to model avian habitat
suitability; \citet{clawges2008} found that lidar derived forest
structural attributes are capable of identifying habitat types
associated with avian species in pine/aspen forests; \citet{goetz2007}
used canopy structural diversity to predict bird species richness,
finding that canopy vertical distribution was the strongest predictor of
species richness. Forest structural metrics have also been used to study
other clades and habitats \citep{davies2014, nelson2005}. Many metrics
derived from lidar remote sensing have been used as local indicators of
biodiversity, including simple metrics such as canopy cover and canopy
height as well as derived metrics including vertical profiles,
aboveground biomass \citep{lefsky1999}. Other second order derived
metrics such as canopy texture, height class distribution, edges, and
patch metrics have also been used to examine habitat and biodiversity at
landscape scales \citep{bergen2009}.

Simple productivity (NDVI, indices) to complex aggregates (DHIs)

@@

Biodiversity has been shown to be linked to both forest structure
\citep{guo2017, gao2014}, and productivity \citep{radeloff2019} at a
range of scales. Research has highlighted that the relationship between
productivity and biodiversity is reciprocal \citep{worm2003}, making
productivity a suitable biodiversity indicator. Within a forest
environment, there are many possible ecosystem structure and function
metrics that can be assessed. Structural variables range in complexity
from simple (canopy cover; canopy height), to more complex (elevation
coefficient of variation; elevation standard deviation) to modelled
(aboveground biomass; basal area), all of which can be assessed using
lidar data \citep{coops2021}. Data on ecosystem functionality ranges
from phenology metrics (date of leaf on/leaf off; length of growing
season) to functional traits (chlorophyll content; leaf area index) to
productivity estimates (GPP; total available energy)
\citep{pettorelli2018}. Advances in satellite remote sensing processing
have allowed 3d forest structure data to be imputed across wide spatial
scales \citep{matasci2018, coops2021} using data fusion approaches
involving collected lidar data and optical/radar data. Other advances in
image compositing have allowed yearly summaries of vegetation
productivity to be calculated at regional to global scales, summarizing
the yearly total, minimum, and variation in energy availability at
similar spatial resolutions \citep{radeloff2019}.

In forested ecosystems, stand structural attributes have also been shown
to influence productivity, with a range of responses dependent on
environmental conditions \citep{ali2019}. Linkages between forest
structure and productivity (namely, vegetation indices) have been
examined for nearly 20 years
\citetext{\citealp[\citet{knyazikhin1998}]{huete2002}; \citealp{myneni1994}}.
While there is significant theoretical and empirical evidence for their
relationship at single time points (within a single image)
\citep{myneni1994}, various relationship directions and shapes have been
found between forest structure and productivity metrics \citep{ali2019}.
These relationships, their shapes, and their strengths have been
attributed to multiple possible hypotheses and have been shown to vary
based on environmental conditions \citep{ali2019}. The relationship
between forest structural diversity metrics and annual productivity
summaries has yet to be fully examined, including the DHIs.

Understanding the complementarity between potential EBVs is an integral
component of their creation \citep{skidmore2021}. Structural equation
modelling and path analyses have been commonly used in ecology to assess
the causal effects behind various hypotheses \citep{fan2016, grace2010},
including the relationship between forest structure, functioning, and
biodiversity \citep{ali2019}. In this paper we seek to untangle the
relationship between two EBVs: forest structure diversity metrics and
yearly summaries of forest productivity. To accomplish this, we assess
this relationship using path analysis to assess the direct and indirect
(as mediated by more complex forest structural diversity metrics)
effects of commonly collected forest structural metrics (canopy height
and canopy cover) on yearly productivity summaries. Further, we use
exploratory structural equation modelling (ESEM) to identify the latent
variables driving the productivity metrics. We run this analysis both
globally, and stratified by the forested ecosystems of British Columbia,
Canada, and compare models using AIC scores and global fit measures
across the ecosystems. The results from this study will assess the
linkages and complementarity between two biodiversity indicators.

@@

We conducted model based sampling across the fifteen forest dominated
ecosystems found within British Columbia {[}BC figure reference, I have
not made this figure yet{]}. Samples were randomly selected within BEC
zones of interest alongside multiple criteria. Each sampled pixel had to
have a forested land cover class (coniferous, deciduous, mixed-wood, or
wetland-treed), and be surrounded by the same land cover class.
Additionally, each pixel had to have a coefficient of variation less
than 0.5 in the Lorey's height and canopy cover forest structural
metrics. A maximum of 3000 samples were sampled in each BEC zone.

Due to variations in ecosystem size, differing numbers of samples were
generated for each ecosystem. To account for this during model
comparison, each ecosystem and land cover sample set was downsampled to
the minimum number of samples for a single BEC zone (in this case, 170).
To meet the normality assumptions of path analysis, all variables were
natural-log transformed and standardized, as per \citet{grace2016}.
Variables containing zeros were natural-log plus one transformed.

@@

\hypertarget{software}{%
\subsection{Software}\label{software}}

The processing for this paper was done in R version 4.2.2
\citep{R-base}. We used \emph{sgsR} \citep{goodbody2023} for sampling,
\emph{terra} \citep{R-terra} for raster analyses, \emph{sf} for vector
analyses \citep{R-sf}, and \emph{lavaan} \citep{lavaan2012} for path
analysis and structural equation modelling.

These focal analyses were preprocessed in R and Python, and randomly
sampled across a masked suitability raster for each BEC zone using the
\emph{sgsR} R package version 1.3.4 \citep{goodbody2023}.

@@

{\centering \includegraphics[width=7in,height=\textheight]{../../outputs/latent_groups.png}

@@

\renewcommand\refname{References}
  \bibliography{bibliography.bib,packages.bib}
